---
output:
  html_document: default
  pdf_document: default
---
# based on an example from http://faculty.washington.edu/kenrice/sisg/SISG-08-06.pdf
---
title: "SimplePerm"
author: "Paul M"
date: "April 5, 2016"
output: html_document
---

Generate two sets of test data: one with an effect ("alt.y"); one with no effect ("null.y"")
```{r}
set.seed(4321)
carrier<-rep(c(0,1), c(100,200))  # Make up an example with 100 carriers and 200 controls
null.y<-rnorm(300)   # our null data - here the value is independent of carrier status
alt.y<-rnorm(300, mean=carrier/2)   # our alternate data - here the mean depends on carrier status
```

Since our data are normally distributed, in this case we know that the test statistic will actually have a t-distribution. So we can do the exact significance test as a benchmark
```{r}
t.test(null.y~carrier, var.equal=TRUE)
t.test(alt.y~carrier, var.equal=TRUE)
```

Now lets work on the permutation test:
```{r}
# Here's the observed difference in mean between the two groups for the null data
null.diff<-mean(null.y[carrier==1])-mean(null.y[carrier==0])
# And here's the observed difference in mean between the two groups for our alternate data
alt.diff<-mean(alt.y[carrier==1])-mean(alt.y[carrier==0])
```

Define a function to conduct a single permutation and report the difference in means for the permuted data
```{r}
one.test<-function(x,y){
  xstar<-sample(x)  # samples without replacement - so this is producing a random permutation of x
  m<-mean(y[xstar==1])-mean(y[xstar==0])  # then calcluate the difference in means between the two groups defined by the permuted x
  return (m)
}
```

Perform permutation test, with 1000 permutations, for each of the null and alt datasets, and then plot results
```{r}
many.truenull <- replicate(1000, one.test(carrier, null.y))
many.falsenull <- replicate(1000, one.test(carrier, alt.y))

# plot histograms of the resulting distribution of difference of means
p1<-mean(abs(many.truenull) > abs(null.diff)) # the empirical p-value (for 2-sided test)
hist(many.truenull,main = c("Null p-value~",p1),breaks=20)
# add a line to it to indicate the 'observed' value
abline(v=null.diff, lwd=2, col="purple")

#text(x=-0.4,y=200,c("p-value=",mean/1000))
p2<-mean(abs(many.falsenull) > abs(alt.diff))  # the empirical p-value
hist(many.falsenull,main = c("Alt p-value~",p2),breaks=20)
abline(v=alt.diff, lwd=2, col="purple")
```

Note that in the case of the alt data, none of the differences in the permuted
datasets exceeded true difference, so all we can really say is that we estimate that p<1/1000.
